---
title: "Understanding mixed effects models through simulating data"
author: 'Lisa DeBruine and Dale Barr'
output: html_document
bibliography: bibliography.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo       = TRUE,
                      fig.path   = "images/",
                      fig.width  = 8,
                      fig.height = 5,
                      fig.align  = "center",
                      out.width  = "100%",
                      warning    = FALSE,
                      message    = FALSE,
                      fig.retina = 2)
```

## Abstract

Experimental designs that sample both subjects and stimuli from a larger population need to account for random effects of both subjects and stimuli using mixed effects models. However, much of this research is analyzed using ANOVA on aggregated responses because researchers are not confident specifying and interpreting mixed effects models. The tutorial will explain how to simulate data with random effects structure and analyse the data using linear mixed effects regression (with the lme4 R package). The focus will be on interpreting the LMER output in light of the simulated parameters and comparing the results to by-items and by-subjects ANOVA.

### Prerequisite knowledge or skills

* Basic familiarity with experimental designs where subjects respond to stimuli
* Basic familiarity with R

### Who will benefit from this tutorial?

Researchers who use experimental designs that need to account for crossed random effects (e.g., designs that sample subjects and stimuli). For example, a large amount of experimental research in face perception or social cognition uses designs that would be better analysed using mixed effects models.

## Generalizing to a population of encounters

Many research questions in psychology and neuroscience are questions about certain types of *events*: What happens when people encounter particular types of stimuli? For example: Do people recognize abstract words faster than concrete words? What impressions do people form about a target person's personality based on their vocal qualities? Can people categorize emotional expressions more quickly on the faces of social in-group members than on the faces of out-group members? How do brains respond to threatening versus non-threatening stimuli? In all of these situations, researchers would like to be able to make general statements about phenomena that go beyond the particular participants and particular stimuli that they happen to have chosen for the specific study. Traditionally, people speak of such designs as having crossed random factors of participants and stimuli, and have discussed the problem as one of simultaneous generalization to both populations. However, it may be more intuitive to think of the problem as wanting to generalize to a single population of events: in particular, to a population of *encounters* between the units from the sampled populations [@barr_2018].

Most analyses using conventional statistical techniques, such as analysis of variance and t-test, commit the fallacy of treating stimuli as fixed rather than random. The problem, and the solutions to the problem, have been known in psycholinguistics for over 50 years [@coleman_1964,@clark_1973], and most psycholinguistic journals require authors to demonstrate generality of findings over stimuli as well as over subjects.  Even so, the quasi-$F$ statistics for ANOVA ($F'$ and min-$F'$) that Clark proposed as a solution were widely recognized as unreasonably conservative [@forster_dickinson_1976], and until fairly recently, most psycholinguists performed separate by-subjects ($F_1$) and by-items analyses ($F_2$), declaring an effect "significant" only if it was significant for both analyses. The $F_1 \times F_2$ approach was widely used, despite the fact that Clark had already shown it to be invalid, since both $F$ statistics have higher than nominal false positives in the presence of a null effect, $F_1$ due to unmodeled stimulus variance, and $F_2$ due to unmodeled subject variance.

Recently, psycholinguists have adopted linear mixed-effects modeling as the standard for the statistical analysis, given numerous advantages over ANOVA, including the ability to simultaneously model subject and stimulus variation, to gracefully deal with missing data or unbalanced designs, and to accommodate arbitrary types of continuous and categorical predictors or response variables [@baayen_davidson_bates_2008,@locker_hoffman_bovaird_2007]. With an appropriately specified model, mixed-effects models yield major improvements in power over quasi-$F$ approaches and avoid the increased false positive rate associated with separate $F_1$ and $F_2$ [@barr_et_al_2013].

Despite mixed-effects modeling becoming the *de facto* standard for analysis in psycholinguistics, the approach has yet to take hold in other areas where stimuli are routinely sampled, even in spite of repeated calls for improved analyses in social psychology [@judd_westfall_kenny_2012] and neuroimaging [@bedny_aguirre_thompson-schill_2007,@westfall_yarkoni_2016]. One of the likely reasons for the limited uptake outside of psycholinguistics is because mixed-effects models expose the analyst to a level of statistical and technical complexity far beyond most researchers' training. While some of this complexity is specific to mixed-effects modeling, some of it is simply hidden away from users of traditional techniques by GUIs and function defaults. The novice mixed modeler is suddenly confronted with the need to make decisions about how to specify categorical predictors, which random effects to include or exclude, which of the statistics in the voluminous output to attend to, and whether and how to re-configure the optimizer function when a convergence error or singularity warning appears.

We see great promise in the mixed-effects approach for improving the generality and thus reproducibility in psychology and related fields, but empathize with the frustration and often, exasperation expressed by many novices when they attempt to grapple with these models in their research. Much of the uncertainty and unease around mixed-effects models comes from using them in situations where the ground truth is unknown. A profitable way to improve understanding and user confidence is through data simulation. Knowing the ground truth allows the user to experiment with various modeling choices and observe their impact on a model's performance.

## Simulating data with crossed random factors

For this tutorial, we will simulate data from a design with crossed random factors of subjects and stimuli. In this hypothetical study, subjects classify the emotional expressions of faces as quickly as possible, and we use their response time as the primary dependent variable. The faces are of two types: either from the subject's in-group or from an out-group; the reader is invited to fill in their favorite social or biological dimension as the distinguishing feature. Our question is whether there is any difference in classification speed across the type of face. 

We will simulate data from 100 raters responding to 50 faces: 25 from the ingroup and 25 from the outgroup. 

### Packages

You will need lme4 [@lme4] and lmerTest [@lmerTest] for mixed effects models (it adds p-values to lme4), tidyverse [@tidyverse] for data wrangling and visualisation, faux [@faux] for simulating data with specified parameters, and broom.mixed [@broom.mixed] for extracting data from mixed effects models. Set a seed at the top of your script so that your script returns the same (random) numbers in this simulation (never set seed inside a function).

```{r pkgs}
library(lmerTest)    # for mixed effects models
library(tidyverse)   # for data wrangling and visualisation
library(faux)        # devtools::install_github("debruine/faux")
library(broom.mixed) # for extracting data from mixed effects models

set.seed(8675309)
```

### Set Study Parameters

Set up your study parameters. In this example, 100 subjects will respond to 50 items; 25 items in each of 2 conditions. For crossed mixed effect models, you can't think of factors as being "within" or "between", but as within- or between-*subjects* **and** within- or between-*items*. In this example, `condition` is a within-subject and between-item factor.

```{r params}
nsubj  <- 100 # number of subjects per between-subject cell
nitem  <- 25  # number of items per between-item cell
```

Then specify the fixed effects. We'll set the overall mean response time (`mu`) to 800 ms and the difference between ingroup and outgroup conditions (`eff`) to 80 ms. 

```{r param-mu-eff}
mu     <- 800 # grand mean
eff    <- 80  # effect of condition
```

Now we specify the variance for the random effects by items and subjects, the correlations among random effects by items and subjects, and the error variance.

** EXPLAIN MORE **

```{r param-var}
# items
iri_sd <- 80  # by-item random intercept sd

# subjects
sri_sd <- 100 # by-subject random intercept sd
srs_sd <- 40  # by-subject random slope sd
rcor <- .2    # correlation between intercept and slope

# error
err_sd <- 200 # residual sd
```

### Simulate Items

We need to create a table listsing each item, its between-item factors, and simulated values for its random effects.

Simulate items using the function `sim_design()` from faux. This function will create a dataset with `n` items per between-cell with the specified means (`mu`), standard deviations (`sd`) and correlations (<code>r</code>). See the [vignette](https://debruine.github.io/faux/articles/sim_design.html) for more details.

Condition is a between-items factor, so we need to include it in the `between` argument. Set `n = nitem` to specify `r nitem` items per condition. There are no within-items factors, so we only need to simulate one random effect: the  intercept. Set `sd = iri_sd` to set the standard deviation for this random effect (the mean defaults to 0). Set `dv = "iri"` to give the DV column that name. Set `id = "item_id"`; we'll use this later to join this information to the table of trials.

```{r sim-items, out.width = "75%", fig.cap="The distribution of random effects for ingroup and outgroup faces."}
items <- faux::sim_design(
  between = list(condition = c("ingroup", "outgroup")),
  n = nitem,
  sd = iri_sd,
  dv = "iri",
  id = "item_id",
  plot = FALSE
)

plot_design(items)
```

We will also set an **effect code** for condition. Since we predict that repsonses to ingroup faces will be faster than outgroup faces, we set ingroup to -0.5 and outgroup to +0.5.

```{r effect-code}
# effect code condition
items$cond <- recode(items$condition, "ingroup" = -0.5, "outgroup" = +0.5)
```


### Simulate Subjects

We also need to create a table listsing each subject, their between-subject factors, and simulated values for their random effects. 

There are no between-subject factors. Condition is within-subjects, so we need to simulate two random effects: the intercept and the slope for the effect of condition. Set the `within` argument in `sim_design()` to a list with one factor (`effect`) that has two levels: `sri` and `srs`. If you set a factor's levels as a named vector, the names (`sri` and `srs`) become the column names in the data table and the values are used in plots created by faux.

Set `n = nsubj` to specify the number of subjects. There are two random effects to specify standard deviation for, so set `sd` using a named vector and set their correlation with <code>r = rcor</code>. Set `dv = "value"`; this will only be used in faux plots. Set `id = "subj_id"`; we'll use this later to join this information to the table of trials.

```{r sim-subjects, out.width = "75%", fig.cap="The distribution of random effects for subjects"}
subjects <- faux::sim_design(
  within = list(effect = c(sri = "By-subject random intercepts", 
                           srs = "By-subject random slopes")), 
  n = nsubj,
  sd = c(sri = sri_sd, srs = srs_sd), 
  r = rcor,
  dv = "value",
  id = "subj_id",
  plot = FALSE
)

plot_design(subjects)
```


### Simulate Trials

Since all subjects respond to all items, we can set up a table of trials by crossing the subject IDs with the item IDs. Each trial has random error associated; we simulate this from a normal distribution with a mean of 0 and SD of `err_sd`.

```{r}
trials <- crossing(subj_id = subjects$subj_id,
                   item_id = items$item_id) %>%
  mutate(err = rnorm(nrow(.), mean = 0, sd = err_sd))
```

Now that we have a table of all trials, we can join the information from our `subjects` and `items` tables using the `subj_id` and `item_id`. Calculate the reponse (`Y`) as the sum of:

* the grand intercept (`mu`), 
* the effect of condition (`eff`) multiplied by the effect-coded condition (`cond`)
* the item-specific random intercept (`iri`), 
* the subject-specific random intercept (`sri`), 
* the subject-specific random slope of condition (`srs`) multiplied by the effect-coded condition (`cond`)
* the random error (`err`)

```{r}
dat_sim <- trials %>%
  inner_join(subjects, "subj_id") %>%
  inner_join(items, "item_id") %>%
  # mutate(Y = mu + iri + sri + (eff+srs)*cond + err) # DALE is below better?
  mutate(Y = mu + (eff*cond) + iri + sri + (srs*cond) + err)
```

### Analyse Data

Now we're ready to analyse our simulated data. The formula for `lmer()` maps onto how we calculated the response above.

```
Y ~ 1 + cond + (1 | item_id) + (1 + cond | subj_id)
```

* `Y` is the response
* `1` is the grand intercept (`mu`), 
* `cond` is the effect of condition (`eff*cond`),
* `1` in `(1 | item_id)` is the item-specific random intercept (`iri`), 
* `1` in `(1 + cond | subj_id)` is the subject-specific random intercept (`sri`), 
* `cond` in `(1 + cond | subj_id)` is the subject-specific random slope of condition (`srs*cond`)

The `lmer()` function takes this formula as its first argument, then the data table. Set `REML = FALSE` to choose the method for estimating variance components. `REML = TRUE` is better when you have fairly unequal cell sizes. **EXPLAIN THIS BETTER**

```{r}
mod_sim <- lmer(Y ~ 1 + cond + (1 | item_id) + (1 + cond | subj_id),
                data = dat_sim, REML = TRUE)
```

```{r echo = FALSE}
srfx <- attr(VarCorr(mod_sim)$subj_id, "stddev") %>% round(2)
irfx <- attr(VarCorr(mod_sim)$item_id, "stddev") %>% round(2)
rc   <- attr(VarCorr(mod_sim)$subj_id, "correlation")[1, 2] %>% round(2)
res  <- attr(VarCorr(mod_sim), "sc") %>% round(2)
ffx  <- fixef(mod_sim) %>% round(2)
```

Use the `summary()` function to view the results. Notice where the parameters you set at the beginning show up in the results.

| variable | explanation                     | simulated value | estimated by model |
|:---------|:--------------------------------|----------------:|-------------------:|
| `mu`     | grand mean                      | `r mu`          | `r ffx[[1]]`       |
| `eff`    | effect of condition             | `r eff`         | `r ffx[[2]]`       |
| `iri_sd` | by-item random intercept SD     | `r iri_sd`      | `r irfx[[1]]`      |
| `sri_sd` | by-subject random intercept SD  | `r sri_sd`      | `r srfx[[1]]`      |
| `srs_sd` | by-subject random slope SD      | `r srs_sd`      | `r srfx[[2]]`      |
| `rcor`   | cor between intercept and slope | `r rcor`        | `r rc`             |
| `err_sd` | residual SD                     | `r err_sd`      | `r res`            |


```{r}
summary(mod_sim, corr = FALSE)
```

You can also use `broom.mixed::tidy()` to output fixed and/or random effects in a tidy table. This is especially useful when you need to combine the output from hundreds of simulations to calculate power.

```{r, results = 'asis'}
broom.mixed::tidy(mod_sim) %>% knitr::kable(digits = 3)
```

### Calculate Power

You can set up a function that takes all of the parameters we set above as arguments. We'll set the to default to the values we used, but you can choose your own defaults. The code below is just all of the code above, condensed a bit.

```{r}
my_sim_func <- function(nsubj  = 100, # number of subjects pre between-subject cell
                        nitem  = 25,  # number of items per between-item cell
                        mu     = 800, # grand mean
                        eff    = 80,  # effect of condition
                        iri_sd = 80,  # by-item random intercept sd (omega_00)
                        sri_sd = 100, # by-subject random intercept sd
                        srs_sd = 40,  # by-subject random slope sd
                        rcor   = .2,  # correlation between intercept and slope
                        err_sd = 200  # residual (standard deviation)
                        ) {
  # simulate items
  items <- faux::sim_design(
    between = list(condition = c("ingroup", "outgroup")),
    n = nitem,
    sd = iri_sd,
    dv = "iri",
    id = "item_id",
    plot = FALSE
  )

  # effect code condition
  items$cond <- recode(items$condition, "ingroup" = -0.5, "outgroup" = 0.5)
  
  # simulate subjects
  subjects <- faux::sim_design(
    within = list(effect = c(sri = "By-subject random intercepts", 
                             srs = "By-subject random slopes")), 
    n = nsubj,
    sd = c(sri = sri_sd, srs = srs_sd), 
    r = rcor,
    dv = "value",
    id = "subj_id",
    plot = FALSE
  )
  
  # simulate trials
  dat_sim <- crossing(subj_id = subjects$subj_id,
                     item_id = items$item_id) %>%
    mutate(err = rnorm(nrow(.), mean = 0, sd = err_sd)) %>%
    inner_join(subjects, "subj_id") %>%
    inner_join(items, "item_id") %>%
    mutate(Y = mu + (eff*cond) + iri + sri + (srs*cond) + err)
  
  mod_sim <- lmer(Y ~ cond + (1 | item_id) + (1 + cond | subj_id),
                dat_sim, REML = FALSE)
  
  broom.mixed::tidy(mod_sim)
}
```

Run the function once with default parameters. 

```{r, results = 'asis'}
my_sim_func() %>% knitr::kable(digits = 3)
```

You can use the `purrr::map_df` function to run the simulation repeatedly and save the results to a data table. This will take a while, so test using just a few reps first, then make sure you save the full results to a CSV file so you can set this code chunk to  not run (`eval = FALSE` in the chunk header) and load from the saved data for the rest of your script in the future.

```{r sim, eval = FALSE}
reps <- 1000
sims <- purrr::map_df(1:reps, ~my_sim_func())
write_csv(sims, "sims.csv")
```

```{r read-sim}
sims <- read_csv("sims.csv")
```

You can use these data to calculate power for each fixed effect or plot the distribution of your fixed or random effects.

```{r sim-fixef-plot, echo = FALSE, fig.cap = "Distribution of fixed effects across 1000 simulations"}
alpha <- 0.05

sumdat <- sims %>% 
  filter(effect == "fixed") %>%
  group_by(term) %>%
  summarise(
    power = paste("power =", mean(p.value < alpha) %>% round(3)),
    xval = mean(estimate) - 2.5*sd(estimate),
    value = mean(estimate)
  )

sims %>%
  filter(effect == "fixed") %>%
  mutate(significant = (p.value < alpha) %>% factor(levels = c(TRUE, FALSE))) %>%
  ggplot() +
  geom_density(aes(estimate, y = ..count.., fill = significant, color = significant), alpha = 0.5) +
  geom_vline(data = sumdat, aes(xintercept = value), color = "grey40", show.legend = FALSE) +
  geom_text(data = sumdat, aes(x = xval, y = 20, label = power), color = "black", show.legend = FALSE) +
  facet_wrap(~term, ncol = 1, scales = "free_x") +
  scale_fill_brewer(palette = "Spectral") + theme_bw()
```



```{r sim-ranef-plot, echo = FALSE, fig.cap = "Distribution of random effects across 1000 simulations"}

sim_stats <- sims %>%
  filter(effect == "ran_pars") %>%
  group_by(group, term) %>%
  summarise(value = mean(estimate))

sims %>%
  filter(effect == "ran_pars") %>%
  ggplot(aes(estimate, color = paste(group, term), fill = paste(group, term))) +
  geom_density(alpha = 0.5, show.legend = FALSE) +
  geom_vline(data = sim_stats, aes(xintercept = value, color = paste(group, term)), show.legend = FALSE) +
  facet_wrap(~group*term, ncol = 3, scales = "free") +
  scale_fill_brewer(palette = "Spectral") + theme_bw()
```








