---
title: 'Appendix 1: Example Code'
subtitle: "Understanding mixed effects models through data simulation"
author: "Lisa M. DeBruine & Dale J. Barr"

output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Appendix 1: Example Code}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 8,
  fig.height = 5,
  out.width = "100%"
)
```


## Simulate data with crossed random factors

To give an overview of the simulation task, we will simulate data from a design with crossed random factors of subjects and stimuli, fit a model to the simulated data, and then recover the parameter values we put in from the model output. 

In this hypothetical study, subjects classify the emotional expressions of faces as quickly as possible, and we use their response time as the primary dependent variable. Let's imagine that the faces are of two types: either from the subject's ingroup or from an outgroup. For simplicity, we further assume that each face appears only once in the stimulus set. The key question is whether there is any difference in classification speed across the type of face.

The important parts of the design are:

* Random factor 1: subjects (associated variables will be prefixed by `T` or `tau`)
* Random factor 2: faces (associated variables will be prefixed by `O` or `omega`)
* Fixed factor 1: category (level = ingroup, outgroup)
    * within subject: subjects see both ingroup and outgroup faces
    * between face: each face is either ingroup or outgroup

### Required software

```{r, message=FALSE}
# load required packages
library("lme4")        # model specification / estimation
library("afex")        # anova and deriving p-values from lmer
library("broom.mixed") # extracting data from model fits 
library("faux")        # generate correlated values
library("tidyverse")   # data wrangling and visualisation

# ensure this script returns the same results on each run
set.seed(8675309)
faux_options(verbose = FALSE)
```

### Establish the data-generating parameters

```{r params-all}
# set all data-generating parameters
beta_0  <- 800 # intercept; i.e., the grand mean
beta_1  <-  50 # slope; i.e, effect of category
omega_0 <-  80 # by-item random intercept sd
tau_0   <- 100 # by-subject random intercept sd
tau_1   <-  40 # by-subject random slope sd
rho     <-  .2 # correlation between intercept and slope
sigma   <- 200 # residual (error) sd
```


### Simulate the sampling process

```{r params}
# set number of subjects and items
n_subj     <- 100 # number of subjects
n_ingroup  <-  25 # number of items in ingroup
n_outgroup <-  25 # number of items in outgroup
```

#### Simulate the sampling of stimulus items

```{r}
# simulate a sample of items
n_items <- n_ingroup + n_outgroup

items <- data.frame(
  item_id = seq_len(n_items),
  category = rep(c("ingroup", "outgroup"), c(n_ingroup, n_outgroup)),
  X_i = rep(c(-0.5, 0.5), c(n_ingroup, n_outgroup)),
  O_0i = rnorm(n = n_items, mean = 0, sd = omega_0)
)
```

#### Simulate the sampling of subjects

You can use one of two methods. The first uses the function `MASS::mvrnorm`, but you have to calculate the variance-covariance matrix yourself. This gets more complicated as you have more variables (e.g., in a design with more than 1 fixed factor).

```{r}
# simulate a sample of subjects

# calculate random intercept / random slope covariance
covar <- rho * tau_0 * tau_1

# put values into variance-covariance matrix
cov_mx  <- matrix(
  c(tau_0^2, covar,
    covar,   tau_1^2),
  nrow = 2, byrow = TRUE)

# generate the by-subject random effects
subject_rfx <- MASS::mvrnorm(n = n_subj,
                             mu = c(T_0s = 0, T_1s = 0),
                             Sigma = cov_mx)

# combine with subject IDs
subjects <- data.frame(subj_id = seq_len(n_subj),
                       subject_rfx)
```

Alternatively, you can use the function `faux::rnorm_multi`, which generates a table of `n` simulated values from a multivariate normal distribution by specifying the means (`mu`) and standard deviations (`sd`) of each variable, plus the correlations (`r`), which can be either a single value (applied to all pairs), a correlation matrix, or a vector of the values in the upper right triangle of the correlation matrix.

```{r}
# simulate a sample of subjects

# sample from a multivariate random distribution 
subjects <- faux::rnorm_multi(
    n = n_subj, 
    mu = 0, # means for random effects are always 0
    sd = c(tau_0, tau_1), # set SDs
    r = rho, # set correlation, see ?rnorm_multi
    varnames = c("T_0s", "T_1s")
  )

# add subject IDs
subjects$subj_id <- seq_len(n_subj)
```

#### Check your values

```{r}
data.frame(
  parameter = c("omega_0", "tau_0", "tau_1", "rho"),
  value = c(omega_0, tau_0, tau_1, rho),
  simulated = c(
    sd(items$O_0i),
    sd(subjects$T_0s),
    sd(subjects$T_1s), 
    cor(subjects$T_0s, subjects$T_1s)
  )
)
```


#### Simulate trials (encounters)

```{r}
# cross subject and item IDs; add an error term
# nrow(.) is the number of rows in the table
trials <- crossing(subjects, items)  %>%
  mutate(e_si = rnorm(nrow(.), mean = 0, sd = sigma))
```

#### Calculate response values

```{r}
dat_sim <- trials %>%
  mutate(RT = beta_0 + T_0s + O_0i + (beta_1 + T_1s) * X_i + e_si) %>%
  select(subj_id, item_id, category, X_i, RT)
```


#### Plot the data

```{r}
ggplot(dat_sim, aes(category, RT, color = category)) +
  # predicted means
  geom_hline(yintercept = (beta_0 - 0.5*beta_1), color = "orange") +
  geom_hline(yintercept = (beta_0 + 0.5*beta_1), color = "dodgerblue") +
  # actual data
  geom_violin(alpha = 0, show.legend = FALSE) +
  stat_summary(fun = mean,geom="crossbar", show.legend = FALSE) +
  scale_color_manual(values = c("orange", "dodgerblue")) +
  ggtitle("Predicted versus simulated values")
```


### Data simulation function

Once you've tested your data generating code above, put it into a function so you can run it repeatedly. We used pipes to combine a few steps at the end.

```{r}
# set up the custom data simulation function
my_sim_data <- function(
  n_subj      = 100,   # number of subjects
  n_ingroup  =  25,   # number of ingroup stimuli
  n_outgroup =  25,   # number of outgroup stimuli
  beta_0     = 800,   # grand mean
  beta_1     =  50,   # effect of category
  omega_0    =  80,   # by-item random intercept sd
  tau_0      = 100,   # by-subject random intercept sd
  tau_1      =  40,   # by-subject random slope sd
  rho        = 0.2,   # correlation between intercept and slope
  sigma      = 200) { # residual (standard deviation)
  
  # simulate a sample of items
  items <- data.frame(
    item_id = seq_len(n_ingroup + n_outgroup),
    category = rep(c("ingroup", "outgroup"), c(n_ingroup, n_outgroup)),
    X_i = rep(c(-0.5, 0.5), c(n_ingroup, n_outgroup)),
    O_0i = rnorm(n = n_ingroup + n_outgroup, mean = 0, sd = omega_0)
  )

  # simulate a sample of subjects
  subjects <- faux::rnorm_multi(
    n = n_subj, mu = 0, sd = c(tau_0, tau_1), r = rho, 
    varnames = c("T_0s", "T_1s")
  )
  subjects$subj_id <- 1:n_subj
  
  # cross subject and item IDs 
  crossing(subjects, items)  %>%
    mutate(
      e_si = rnorm(nrow(.), mean = 0, sd = sigma),
      RT = beta_0 + T_0s + O_0i + (beta_1 + T_1s) * X_i + e_si
    ) %>%
    select(subj_id, item_id, category, X_i, RT)
}
```


## Analyze the simulated data

```{r}
# fit a linear mixed-effects model to data
mod_sim <- lmer(RT ~ 1 + X_i + (1 | item_id) + (1 + X_i | subj_id),
                data = dat_sim)

summary(mod_sim, corr = FALSE)
```

Use `broom.mixed::tidy(mod_sim)` to get a tidy table of the results. Below, we added column "parameter" and "value", so you can compare the estimate from the model to the parameters you used to simulate the data. 

```{r, echo = FALSE}
# get a tidy table of results
broom.mixed::tidy(mod_sim) %>% 
  mutate_if(is.numeric, round, 3) %>%
  mutate(
    parameter = c("beta_0", "beta_1", "omega_0", "tau_0", "tau_1", "rho", "sigma"),
    value = c(beta_0, beta_1, omega_0, tau_0, tau_1, rho, sigma),
  ) %>%
  select(1:3, 9, 10, 4:8) %>%
  knitr::kable()
```


## Calculate Power

You can wrap up the data generating function and the analysis code in a new function (`single_run`) that returns a tidy table of the analysis results, and optionally saves this info to a file if you set a filename.

```{r}
# set up the power function
single_run <- function(filename = NULL, ...) {
  # ... is a shortcut that forwards any additional arguments to my_sim_data()
  dat_sim <- my_sim_data(...)
  mod_sim <- lmer(RT ~ X_i + (1 | item_id) + (1 + X_i | subj_id),
                dat_sim)
  
  sim_results <- broom.mixed::tidy(mod_sim)
  
  # append the results to a file if filename is set
  if (!is.null(filename)) {
    append <- file.exists(filename) # append if the file exists
    write_csv(sim_results, filename, append = append)
  }
  
  # return the tidy table
  sim_results
}
```


```{r}
# run one model with default parameters
single_run()
```


```{r}
# run one model with new parameters
single_run(n_ingroup = 50, n_outgroup = 45, beta_1 = 20)
```

To get an accurate estimation of power, you need to run the simulation many times. We use 100 here as an example, but your results are more accurate the more replications you run. This will depend on the specifics of your analysis, but we recommend at least 1000 replications.

```{r}
filename <- "sims/sims.csv" # change for new analyses
if (!file.exists(filename)) {
  # run simulations and save to a file
  reps <- 100
  sims <- purrr::map_df(1:reps, ~single_run(filename))
}

# read saved simulation data
sims <- read_csv(filename)
```

```{r}
# calculate mean estimates and power for specified alpha
alpha <- 0.05

sims %>% 
  filter(effect == "fixed") %>%
  group_by(term) %>%
  summarise(
    mean_estimate = mean(estimate),
    mean_se = mean(std.error),
    power = mean(p.value < alpha),
    .groups = "drop"
  )
```


## Compare to ANOVA

One way many researchers would normally analyse data like this is by averaging each subject's reaction times across the ingroup and outgroup stimuli and compare them using a paired-samples t-test or ANOVA (which is formally equivalent). Here, we use `afex::aov_ez` to analyse a version of our dataset that is aggregated by subject.

```{r}
# aggregate by subject and analyze with ANOVA
dat_subj <- dat_sim %>%
  group_by(subj_id, category, X_i) %>%
  summarise(RT = mean(RT), .groups = "drop")

afex::aov_ez(
  id = "subj_id",
  dv = "RT",
  within = "category",
  data = dat_subj
)
```

Alternatively, you could aggregate by item, averaging all subjects' scores for each item.

```{r}
# aggregate by item and analyze with ANOVA
dat_item <-  dat_sim %>%
  group_by(item_id, category, X_i) %>%
  summarise(RT = mean(RT), .groups = "drop")

afex::aov_ez(
  id = "item_id",
  dv = "RT",
  between = "category",
  data = dat_item
)
```


We can create a power analysis function that simulates data using our data-generating process from `my_sim_data()`, creates these two aggregated datasets, and analyses them with ANOVA. We'll just return the p-values for the effect of category as we can calculate power as the percentage of these simulations that reject the null hypothesis.

```{r}
# power function for ANOVA
my_anova_power <- function(...) {
  dat_sim <- my_sim_data(...)
  
  dat_subj <-  dat_sim %>%
    group_by(subj_id, category, X_i) %>%
    summarise(RT = mean(RT), .groups = "drop")
  
  dat_item <-  dat_sim %>%
    group_by(item_id, category, X_i) %>%
    summarise(RT = mean(RT), .groups = "drop")

  a_subj <- afex::aov_ez(id = "subj_id",
                         dv = "RT",
                         within = "category",
                         data = dat_subj)
  suppressMessages(
    # check contrasts message is annoying
    a_item <- afex::aov_ez(
      id = "item_id",
      dv = "RT",
      between = "category",
      data = dat_item
    )
  )
  
  list(
    "subj" = a_subj$anova_table$`Pr(>F)`,
    "item" = a_item$anova_table$`Pr(>F)`
  )
}
```

Run this function with the default parameters to determine the power each analysis has to detect an effect of category of `r beta_1` ms.

```{r anova-power}
# run simulations and calculate power 
reps <- 100
anova_sims <- purrr::map_df(1:reps, ~my_anova_power())

alpha <- 0.05
power_subj <- mean(anova_sims$subj < alpha)
power_item <- mean(anova_sims$item < alpha)
```

The by-subjects ANOVA has power of `r power_subj`, while the by-items ANOVA has power of `r power_item`. This isn't simply a consequence of within versus between design or the number of subjects versus items, but rather a consequence of the inflated false positive rate of some aggregated analyses.

Set the effect of category to 0 to calculate the false positive rate. This is the probability of concluding there is an effect when there is no actual effect in your population.

```{r}
# run simulations and calculate the false positive rate 
reps <- 100
anova_fp <- purrr::map_df(1:reps, ~my_anova_power(beta_1 = 0))

false_pos_subj <- mean(anova_fp$subj < alpha)
false_pos_item <- mean(anova_fp$item < alpha)
```

Ideally, your false positive rate will be equal to alpha, which we set here at `r alpha`. The by-subject aggregated analysis has a massively inflated false positive rate of `r false_pos_subj`, while the by-item aggregated analysis has a closer-to-nominal false positive rate of `r false_pos_item`. This is not a mistake, but a consequence of averaging items and analysing a between-item factor. Indeed, this problem with false positives is one of the most compelling reasons to analyze cross-classified data using mixed effects models.