---
title: 'Appendix 1b: Example Code (no tidyverse)'
subtitle: "Understanding mixed effects models through data simulation"
author: "Lisa M. DeBruine & Dale J. Barr"

output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Appendix 1b: Example Code (no tidyverse)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE
)
```

This script provides equivalent code to the version using pipes and functions from the tidyverse. While this code may be of interest to people who want to learn more about the computations that generate correlated data, it will be more difficult to extend to more complicated designs.

## Simulating data with crossed random factors

In this hypothetical study, subjects classify the emotional expressions of faces as quickly as possible, and we use their response time as the primary dependent variable. Let's imagine that the faces are of two types: either from the subject's ingroup or from an outgroup. For simplicity, we further assume that each face appears only once in the stimulus set. The key question is whether there is any difference in classification speed across the type of face.

The important parts of the design are:

* Random factor 1: subjects (associated variables will be prefixed by T, $\tau$, or `tau`)
* Random factor 2: faces (associated variables will be prefixed by O, $\omega$, or `omega`)
* Fixed factor 1: category (level = ingroup, outgroup)
    * within subject: subjects see both ingroup and outgroup faces
    * between face: faces are either ingroup or outgroup

### Required software

```{r, message=FALSE}
# load required packages
library("lme4")        # model specification / estimation
library("afex")        # anova and deriving p-values from lmer
library("broom.mixed") # extracting data from model fits 
library("MASS")        # generate correlated values

# ensure this script returns the same results on each run
set.seed(8675309)
```

### Establishing the data-generating parameters

```{r params-all}
# set all data-generating parameters
b0      <- 800 # intercept; i.e., the grand mean
b1      <-  50 # slope; i.e, effect of category
tau_0   <- 100 # by-subject random intercept sd
tau_1   <-  40 # by-subject random slope sd
tau_r   <-  .2 # correlation between intercept and slope
omega_0 <-  80 # by-item random intercept sd
sigma   <- 200 # residual (error) sd
```


### Simulating the sampling process

```{r params}
# set number of subjects and items
tau_n   <- 100 # number of subjects
omega_n <-  25  # number of items per category
```

#### Simulate the sampling of stimulus items

```{r}
# simulate a sample of items
items <- data.frame(
  item_id = 1:(omega_n*2),
  category = rep(c("ingroup", "outgroup"), omega_n),
  cat = rep(c(-0.5, +0.5), omega_n), # effect-code category
  O0i = rnorm(omega_n*2, 0, omega_0)
)
```


#### Simulate the sampling of subjects

```{r}
# simulate a sample of subjects

# make the correlation matrix
cormat <- matrix(c(    1, tau_r,
                   tau_r,     1), 
             nrow = 2, byrow = TRUE) 

# make a corresponding matrix of the variance 
# (multiply the SDs for each cell)
varmat <- matrix(c(tau_0 * tau_0, tau_0 * tau_1,
                   tau_0 * tau_1, tau_1 * tau_1), 
             nrow = 2, byrow = TRUE) 

# create correlated variables with the specified parameters
S <- MASS::mvrnorm(n = tau_n, mu = c(0, 0), 
                   Sigma = cormat * varmat)

subjects <- data.frame(
  subj_id = 1:tau_n,
  T0s = S[, 1],
  T1s = S[, 2]
)
```

#### Check your values

```{r}
data.frame(
  parameter = c("omega_0", "tau_0", "tau_1", "tau_r"),
  value = c(omega_0, tau_0, tau_1, tau_r),
  simulated = c(
    sd(items$O0i),
    sd(subjects$T0s),
    sd(subjects$T1s), 
    cor(subjects$T0s, subjects$T1s)
  )
)
```


#### Simulate trials (encounters)

```{r}
# cross subject and item IDs; add an error term
trials <- expand.grid(subj_id = subjects$subj_id,
                      item_id = items$item_id)

trials$sigma = rnorm(nrow(trials), mean = 0, sd = sigma)

# join subject and item tables
joined <- merge(trials, subjects, by = "subj_id")
dat_sim <- merge(joined, items, by = "item_id")
```

#### Calculate the response values


```{r}
# calculate the response variable
dat_sim$RT <- b0 + dat_sim$O0i + dat_sim$T0s + 
  (b1 + dat_sim$T1s) * dat_sim$cat + dat_sim$sigma

```

#### Plot the data

```{r}
par(mfrow=c(2, 1))
ingroup <- dat_sim$RT[which(dat_sim$category == "ingroup")]
hist(ingroup)
outgroup <- dat_sim$RT[which(dat_sim$category == "outgroup")]
hist(outgroup)
```

### Data simulation function

Once you've tested your data generating code above, put it into a function so you can run it repeatedly. We combined a few steps in calculating Sigma for the correlated values.

```{r}
# set up the custom data simulation function
my_sim_data <- function(
  tau_n   = 100, # number of subjects
  omega_n =  25,  # number of items per category
  b0      = 800, # grand mean
  b1      =  50, # effect of category
  omega_0 =  80, # by-item random intercept sd
  tau_0   = 100, # by-subject random intercept sd
  tau_1   =  40, # by-subject random slope sd
  tau_r   = 0.2, # correlation between intercept and slope
  sigma   = 200  # residual (standard deviation)
  ) {
  
  # simulate items
  items <- data.frame(
    item_id = 1:(omega_n*2),
    category = rep(c("ingroup", "outgroup"), omega_n),
    cat = rep(c(-0.5, +0.5), omega_n), # effect-code category
    O0i = rnorm(omega_n*2, 0, omega_0)
  )
  
  # simulate subjects
  Sigma <- matrix(c(tau_0^2, tau_0 * tau_1 * tau_r,
                    tau_0 * tau_1 * tau_r, tau_1^2), 
               nrow = 2, byrow = TRUE) 
  S <- MASS::mvrnorm(tau_n, c(0, 0), Sigma)
  
  subjects <- data.frame(
    subj_id = 1:tau_n,
    T0s = S[, 1],
    T1s = S[, 2]
  )
  
  # cross subject and item IDs; add an error term
  trials <- expand.grid(subj_id = subjects$subj_id,
                        item_id = items$item_id)
  
  trials$sigma = rnorm(nrow(trials), mean = 0, sd = sigma)
  
  # join subject and item tables
  joined <- merge(trials, subjects, by = "subj_id")
  dat_sim <- merge(joined, items, by = "item_id")
  
  # calculate the response variable
  dat_sim$RT <- b0 + dat_sim$O0i + dat_sim$T0s + 
    (b1 + dat_sim$T1s) * dat_sim$cat + dat_sim$sigma
  
  dat_sim
}
```

## Analyzing the simulated data

```{r}
# fit a linear mixed-effects model to data
mod_sim <- lmer(RT ~ 1 + cat + (1 | item_id) + (1 + cat | subj_id),
                data = dat_sim, REML = TRUE)

summary(mod_sim, corr = FALSE)
```


```{r}
# get a tidy table of results
tbl <- broom.mixed::tidy(mod_sim)
tbl$sim <- c(b0, b1, tau_0, tau_r, tau_1, omega_0, sigma)
tbl[, c(1:3, 9, 4:8)]
```


## Calculate Power

```{r}
# set up the power function
my_lmer_power <- function(filename = NULL, ...) {
  # ... is a shortcut that forwards any arguments to my_sim_data()
  dat_sim <- my_sim_data(...)
  mod_sim <- lmer(RT ~ cat + (1 | item_id) + (1 + cat | subj_id),
                dat_sim, REML = FALSE)
  
  sim_results <- broom.mixed::tidy(mod_sim)
  
  # append the results to a file if filename is set
  if (!is.null(filename)) {
    append <- file.exists(filename) # append if the file exists
    write.table(sim_results, filename, sep = ",",
                row.names = FALSE,
                append = append, 
                col.names = !append)
  }
  
  # return the tidy table
  sim_results
}
```


```{r}
# run one model with default parameters
my_lmer_power()
```


```{r}
# run one model with new parameters
my_lmer_power(omega_n = c(ingroup = 50, outgroup = 50), b1 = 20)
```


```{r, eval = FALSE}
# run simulations
reps <- 100
sims <- replicate(reps, my_lmer_power(), simplify = FALSE)
sims <- lapply(sims, as.data.frame)
sims <- do.call("rbind", sims)
```

Alternatively, you can save the simulations to a file and read from that file in subsequent runs.
```{r}
filename <- "sims/sims_base.csv" # change for new analyses
if (!file.exists(filename)) {
  reps <- 100
  trash <- replicate(reps, my_lmer_power(filename))
}

# read saved simulation data
sims <- read.csv(filename)
```


```{r}
# calculate mean estimates and power for specified alpha
alpha <- 0.05

fcat <- sims[sims$effect == "fixed" & sims$term == "cat", ]

power <- mean(fcat$p.value < alpha)
mean_estimate <- mean(fcat$estimate)
mean_se <- mean(fcat$std.error)
```

The power is `r power` with a mean estimate for the effect of category of `r mean_estimate` and a mean standard error of `r mean_se`.

## Comparison to ANOVA

One way many researchers would normally analyse data like this is by averaging each subject's reaction times across the ingroup and outgroup stimuli and compare them using a paired-samples t-test or ANOVA (which is formally equivalent). Here, we use `afex::aov_ez` to analyse a version of our dataset that is aggregated by subject.

```{r}
# aggregate by subject and analyze with ANOVA
dat_subj <- with(dat_sim, aggregate(RT, list(subj_id, category, cat), mean))
names(dat_subj) <- c("subj_id", "category", "cat", "RT")

afex::aov_ez(
  id = "subj_id",
  dv = "RT",
  within = "category",
  data = dat_subj
)
```

Alternatively, you could aggregate by item, averaging all subjects' scores for each item.

```{r, message=FALSE}
# aggregate by item and analyze with ANOVA
dat_item <- with(dat_sim, aggregate(RT, list(item_id, category, cat), mean))
names(dat_item) <- c("item_id", "category", "cat", "RT")

afex::aov_ez(
  id = "item_id",
  dv = "RT",
  between = "category",
  data = dat_item
)
```

We can create a power analysis function that simulates data using our data-generating process from `my_sim_data()`, creates these two aggregated datasets, and analyses them with ANOVA. We'll just return the p-values for the effect of category as we can calculate power as the percentage of these simulations that reject the null hypothesis.

```{r}
# power function for ANOVA
my_anova_power <- function(...) {
  dat_sim <- my_sim_data(...)
  
  dat_subj <- with(dat_sim, aggregate(RT, list(subj_id, category, cat), mean))
  names(dat_subj) <- c("subj_id", "category", "cat", "RT")
  
  dat_item <- with(dat_sim, aggregate(RT, list(item_id, category, cat), mean))
  names(dat_item) <- c("item_id", "category", "cat", "RT")

  a_subj <- afex::aov_ez(id = "subj_id",
                         dv = "RT",
                         within = "category",
                         data = dat_subj)
  suppressMessages(
    # check contrasts message is annoying
    a_item <- afex::aov_ez(
      id = "item_id",
      dv = "RT",
      between = "category",
      data = dat_item
    )
  )
  
  list(
    "subj" = a_subj$anova_table$`Pr(>F)`,
    "item" = a_item$anova_table$`Pr(>F)`
  )
}
```

Run this function with the default parameters to determine the power each analysis has to detect an effect of category of `r b1` ms.

```{r}
# run simulations and calculate power 
reps <- 100
anova_sims <- replicate(reps, my_anova_power(), simplify = FALSE)
anova_sims <- lapply(anova_sims, as.data.frame)
anova_sims <- do.call("rbind", anova_sims)

alpha <- 0.05
power_subj <- mean(anova_sims$subj < alpha)
power_item <- mean(anova_sims$item < alpha)

```

The by-subjects ANOVA has power of `r power_subj`, while the by-items ANOVA has power of `r power_item`. This isn't simply a consequence of within versus between design or the number of subjects versus items, but rather a consequence of the inflated false positive rate of some aggregated analyses.

Set the effect of category to 0 to calculate the false positive rate. This is the probability of concluding there is an effect when there is no actual effect in your population.

```{r}
# run simulations and calculate the false positive rate 
reps <- 100
anova_fp <- replicate(reps, my_anova_power(b1 = 0), simplify = FALSE)
anova_fp <- lapply(anova_fp, as.data.frame)
anova_fp <- do.call("rbind", anova_fp)

false_pos_subj <- mean(anova_fp$subj < alpha)
false_pos_item <- mean(anova_fp$item < alpha)

false_pos_subj
false_pos_item
```

Ideally, your false positive rate will be equal to alpha, which we set here at `r alpha`. The by-subject aggregated analysis has a massively inflated false positive rate of `r false_pos_subj`, while the by-item aggregated analysis has a closer-to-nominal false positive rate of `r false_pos_item`. This is not a mistake, but a consequence of averaging items and analysing a between-item factor. Indeed, this problem with false positives is one of the most compelling reasons to analyze cross-classified data using mixed effects models.