---
title: 'Appendix 3b: Extended Binomial Example'
subtitle: "Understanding mixed effects models through data simulation"
author: "Lisa M. DeBruine & Dale J. Barr"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Appendix 3b: Extended Binomial Example}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width  = 8,
  fig.height = 5,
  out.width  = "100%"
)
```

[Download the .Rmd for this example](https://github.com/debruine/lmem_sim/blob/master/vignettes/appendix3b_extended_binomial.Rmd)

## Simulating binomial data with crossed random factors

To give an overview of the simulation task, we will simulate data from a design with crossed random factors of subjects and stimuli, fit a model to the simulated data, and then try to recover the parameter values we put in from the output. In this hypothetical study, subjects classify the emotional expressions of faces as quickly as possible, and we use accuracy (correct/incorrect) as the primary dependent variable. The faces are of two types: either from the subject's ingroup or from an outgroup. For simplicity, we further assume that each face appears only once in the stimulus set. The key question is whether there is any difference in classification accuracy across the type of face.

The important parts of the design are:

* Random factor 1: subjects 
* Random factor 2: faces
* Fixed factor 1: expression (level = angry, happy)
    * within subject: subjects see both angry and happy faces
    * within face: same faces are both angry and happy
* Fixed factor 2: category (level = ingroup, outgroup)
    * within subject: subjects see both ingroup and outgroup faces
    * between face: faces are either ingroup or outgroup

### Required software

```{r, message=FALSE}
# load required packages
library("lme4")        # model specification / estimation
library("afex")        # anova and deriving p-values from lmer
library("broom.mixed") # extracting data from model fits 
library("faux")        # data simulation
library("tidyverse")   # data wrangling and visualisation
# ensure this script returns the same results on each run
set.seed(8675309)
faux_options(verbose = FALSE)
```

This example presents an extended simulation for a binomial logistic mixed regression. Where response accuracy is measured in terms of probability, the regression needs to work with a link function that uses the logit, which does not have the problems of being bounded by 0 and 1. The function below is used later to convert between probability and the logit function of probability. 

```{r fig.width = 8, fig.height = 3}
logit <- function(x) { log(x / (1 - x)) }
inv_logit <- function(x) { 1 / (1 + exp(-x)) }
data.frame(
  prob = seq(0,1,.01)
) %>%
  mutate(logit = logit(prob)) %>%
  ggplot(aes(prob, logit)) +
  geom_point()
```


## 2ww*2wb design

In this example, 100 subjects will respond twice (for happy and angry expressions) to 50 items; 25 items in each of 2 categories. In this example, `expression` is a within-subject and within-item factor and `category` is a within-subject and between-item factor.

We use the following prefixes to designate model parameters and sampled values: 
* `beta_*`: fixed effect parameters
* `subj_*`: random effect parameters associated with subjects
* `item_*`: random effect parameters associated with items
* `X_*`: effect-coded predictor
* `S_*`: sampled values for subject random effects
* `I_*`: sampled values for item random effects

In previous tutorials, we used numbers to designate fixed effects, but here we will use letter abbreviations to make things clearer:

* `*_0`: intercept
* `*_e`: expression
* `*_c`: category
* `*_ec`: expression * category

Other terms:

* `*_rho`: correlations; a vector of the upper right triangle for the correlation matrix for that group's random effects
* `n_*`: sample size

Betas are on logit scale; later we use inv_logit to convert to probabilities.
The defaults for the makebinary function below are for null effects.  
When simulating realistic data, we will call the function using beta values computed using expected probabilities for combinations of independent variables. These overwrite the defaults specified here.
All SDs, representing variation in item and subject-specific effects, are set to one.

```{r simulatedata}
ext_bin_data <- function(
  n_subj     = 30, # number of subjects - modified to run faster
  n_ingroup  =  25,  # number of faces in ingroup
  n_outgroup =  25,  # number of faces in outgroup
  beta_0     =   0, # grand mean - inv_logit(0)=.5
  beta_e     =   0, # main effect of expression
  beta_c     =   0, # main effect of category
  beta_ec    =   0, # interaction between category and expression
  item_0    =   1, # by-item random intercept sd
  item_e     =   1, # by-item random slope for exp
  item_rho   =   0, # by-item random effect correlation
  subj_0     =   1, # by-subject random intercept sd
  subj_e     =   1, # by-subject random slope sd for exp
  subj_c     =   1, # by-subject random slope sd for category
  subj_ec    =   1, # by-subject random slope sd for category*exp
  # by-subject random effect correlations
  subj_rho   = c(0, 0, 0, # subj_0  * subj_e, subj_c, subj_ec
                    0, 0, # subj_e  *         subj_c, subj_ec
                       0) # subj_c  *                 subj_ec
) {
  # simulate items;  separate item ID for each item; simulating each item's individual effect for intercept (I_0) and slope
  items <- faux::rnorm_multi(
    n = n_ingroup + n_outgroup,
    mu = 0, 
    sd = c(item_0, item_e),
    r = item_rho,
    varnames = c("I_0", "I_e")
  ) %>%
    mutate(item_id = faux::make_id(nrow(.), "I"),
           category = rep(c("ingroup", "outgroup"), 
                           c(n_ingroup, n_outgroup)))
  
   # simulate subjects: separate subject ID for each subject; simulating each subject's individual effect for intercept (I_0) and slope for each factor and the interaction.
  subjects <- faux::rnorm_multi(
    n = n_subj,
    mu = 0,
    sd = c(subj_0, subj_e, subj_c, subj_ec), 
    r = subj_rho,
    varnames = c("S_0", "S_e", "S_c", "S_ec")
  ) %>%
    mutate(subj_id = faux::make_id(nrow(.), "S"))
  # simulate trials
  # simulate trials
  crossing(subjects, items,
    expression = factor(c("happy", "angry"), ordered = TRUE)
  ) %>%
    mutate(
      # effect code the two fixed factors
      X_e = recode(expression, "happy" = -0.5, "angry" = 0.5),
      X_c = recode(category, "ingroup" = -0.5, "outgroup" = +0.5),
      # add together fixed and random effects for each effect
      B_0  = beta_0  + S_0 + I_0,
      B_e  = beta_e  + S_e + I_e,
      B_c  = beta_c  + S_c,
      B_ec = beta_ec + S_ec,
      # calculate gaussian effect
      Y = B_0 + (B_e*X_e) + (B_c*X_c) + (B_ec*X_e*X_c),
      pr = inv_logit(Y), # transform to probability of getting 1
      Y_bin = rbinom(nrow(.), 1, pr) # sample from bernoulli distribution, ie use pr to set observed binary response to 0 or 1, so if pr = .5, then 0 and 1 equally likely
    ) %>%
    select(subj_id, item_id, expression, category, X_e, X_c, Y, Y_bin)
}
```

```{r showsimdat}
dat_sim <- ext_bin_data()
head(dat_sim)

```
Can compare single run output with the parameters that were set.

```{r singlerun_function}
single_run2 <- function(filename = NULL, ...) {
  dat_sim <- ext_bin_data(...)
  mod_sim <- glmer(Y_bin ~ 1 + X_e*X_c + 
                     (1 + X_e | item_id) + 
                     (1 + X_e*X_c | subj_id),
                data = dat_sim, family = "binomial")
  
  sim_results <- broom.mixed::tidy(mod_sim)
  
  # append the results to a file if filename is set
  if (!is.null(filename)) {
    append <- file.exists(filename) # append if the file exists
    write_csv(sim_results, filename, append = append)
  }
  
  # return the tidy table
  sim_results
}
```



The following function calculates the betas (on logit scale) from mean probability correct for each combination of factors. 
The values entered are the mean probability correct for each combination of effects.
Default is all are 50% correct

You will need to figure out a custom function for each design to do this, or estimate fixed effect parameters from analysis of pilot data.

```{r prob2param}
prob2param <- function(A1B1 = .5,
                       A1B2 = .5,
                       A2B1 = .5,
                       A2B2 = .5) {
  ai <- logit(A1B1)
  ao <- logit(A1B2)
  hi <- logit(A2B1)
  ho <- logit(A2B2)
  list(
    beta_0 = mean(c(ai,ao,hi,ho)), # grand mean
    beta_e = (ao+ai) - (ho+hi), # angry - happy
    beta_c = (ao+ho) - (ai+hi), # outgroup - ingroup
    beta_ec = (ao-ai) - (ho-hi) # angry o-i diff - happy o-i diff
  )
}
```

Demonstrate a single run of the function.

```{r singlerundemo}
#first we specify proportions correct for the 4 combinations. These are just estimated values to show how it works

#to check false positive rate of analysis, set all these to same value (e.g. .5)
A1B1 = .5 #angry_ingroup
A1B2 = .6 #angry_outgroup
A2B1 = .4 #happy_ingroup
A2B2 = .6 #happy_outgroup
b = prob2param(A1B1, A1B2,A2B1,A2B2) 
print('beta values are: ')
b

# e.g. with values set to .5,.6,.4,.5 we get beta_0 and beta_ec of zero (no interaction)
# and beta_e and beta_c of .81 (same size effect of angry/happy and ingroup/outgroup)

# We now do a single run of analysis on simulated data. This is slow - we show how long one run takes using the timer.
ptm <- proc.time() #start clock
sims <- single_run2(
    beta_0 = b$beta_0,
    beta_e = b$beta_e,
    beta_c = b$beta_c,
    beta_ec = b$beta_ec)
# Stop the clock
proc.time() - ptm
sims
```

```{r runsims}
# We now run many simulations and save to a file on each rep

reps <- 10
#we make a filename indicating this specific set of estimated values
filename <- paste0("sims/binomial_",A1B1,"_",A1B2,"_",A2B1,"_",A2B2,"_rep",reps,".csv")
if (!file.exists(filename)) {
  # run simulations and save to a file
  sims <- purrr::map_df(1:reps, ~single_run2(
    filename = filename,
    beta_0 = b$beta_0,
    beta_e = b$beta_e,
    beta_c = b$beta_c,
    beta_ec = b$beta_ec)
  )
}
# read saved simulation data
sims <- read_csv(filename)
```


### Calculate mean estimates

```{r meanests}
est <- sims %>%
  filter(effect == "fixed") %>%
  group_by(term) %>%
  summarise(
    mean_estimate = mean(estimate),
    .groups = "drop"
  )
est %>% mutate(
  parameter = c("beta_0", "beta_c", "beta_e", "beta_ec"),
  value = c(b$beta_0, b$beta_c, b$beta_e, b$beta_ec)
) %>%
  mutate_if(is.numeric, round, 4)

#value is the true value of beta derived from the probabilities we specified
#estimate is the mean estimate of that value in the simulations
```

### Calculate probabilities

Sum estimates for each cell and use inverse logit transform to recover probabilities.

```{r}
int <- est[[1,2]]
cat <- est[[2,2]]
exp <- est[[3,2]]
cat_exp <- est[[4,2]]
data.frame(
  angry_outgroup = inv_logit(int + .25*cat + .25*exp + .25*cat_exp),
  angry_ingroup  = inv_logit(int - .25*cat + .25*exp - .25*cat_exp),
  happy_outgroup = inv_logit(int + .25*cat - .25*exp - .25*cat_exp),
  happy_ingroup  = inv_logit(int - .25*cat - .25*exp + .25*cat_exp)
) %>%
  gather(key, val, 1:4) %>%
  separate(key, c("exp", "cat")) %>%
  spread(cat, val) %>%
  mutate_if(is.numeric, round, 2)

#Can compare with specified values of .5.,.6, .4, .5 in runsims. 
```

## Power analysis

```{r power}

# calculate mean power for specified alpha
alpha <- 0.05
sims %>% 
  filter(effect == "fixed") %>%
  dplyr::group_by(term) %>%
  dplyr::summarise(
    mean_estimate = mean(estimate),
    mean_se = mean(std.error),
    power = mean(p.value < alpha),
    .groups = "drop"
  )

# Note use of dplyr:: with the tidyverse commands group_by and summarise. This avoids problems that can arise from conflicts of these names with other packages.


```






